Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-23 WSSSPE1_report/WSSSPE1_summary_report.tex
Shorten title.
fiddling.

\title[]{Overview \& Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
%\note{\scriptsize Authors, please add/check middle initials and affiliations}

@@

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
This document summarizes the First Workshop on Sustainable Software for Science:
Practice and Experiences (WSSSPE1). The content is based on
a set of collaborative notes taken during the workshop,
including discussion about the keynote presentations, three themed panels that
discussed papers accepted at the workshop`, other discussion, cross-cutting
issues, use cases, and conclusions. It joins a nascent literature seeking to
understand what drives software work in science and how the reward systems of
science function, thereby shaping the types of software projects undertaken, including the
extent to which developers are motivated to build software for the long-term,
for the use of others, and whether to work collaboratively or separately.
Unique perspectives were captured about issues such as comprehensive documentation,
development and deployment practices; software licenses; career paths
for developers; identifiers for software through a
mechanism such as a Digital Object Identifier or publication of a ``software
paper''; and evidence of software contribution and impact, recorded in online
systems such as source code repositories like GitHub. For truly sustainable
software, there should be no endpoint, as the software products continue to be
used and useful beyond a single institution, grant, and developer or
development team.

@@

Another problem Bourne experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of software sustainability through public-private partnerships. He
noted that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding
requirement even with a vibrant community --- goodwill only goes so
far if one is being paid to do something else.  He talked about grant
schemes of relevance in the U.S., particularly with regard to
technology transfer~\cite{sbir-web, fased-web}.

@@

The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon Smith of
GitHub. Smith's talk started with an example from his own past in data
reduction in Astronomy, where he needed to remove interfering effects from the
experimental apparatuses. He built a ``bad pixel mask,'' and realized that
while it was persistent, there was no way or practice of sharing these data
among the scientists. Consequently many researchers repeated the same
calculations. Smith estimated that 13 person-years were wasted by this
repetition.

@@

C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor,
argued that with regard to reusable software, ``we should just start
doing it.'' In this regard Smith replied that documentation should be
``treated as a first class entity.''  He noted that the open source
community has excellent cultures of code reuse, for example,
RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
CPAN~\cite{cpan-web}, where there is effectively low-friction
collaboration through the use of repositories. This has not happened
in highly numerical, compiled language scientific software.  An
exception he cited as a good example of scientific projects using
GitHub is the EMCEE Markov Chain Monte Carlo project~\cite{emcee-web}
by Dan Foreman-Mackey and contributors.

>>>>

\title[]{Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
%\note{\scriptsize Authors, please add/check middle initials and affiliations}

@@

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
This document summarizes the First Workshop on Sustainable Software for Science:
Practice and Experiences (WSSSPE1). The content is based on
a set of collaborative notes taken during the workshop,
including discussion about the keynote presentations, three themed panels that
discussed papers accepted at the workshop`, other discussion, cross-cutting
issues, use cases, and conclusions. It joins a nascent literature seeking to
understand what drives software work in science and how the reward systems of
science function, thereby shaping the types of software projects undertaken, including the
extent to which developers are motivated to build software for the long-term,
for the use of others, and whether to work collaboratively or separately.
Unique perspectives were captured about issues such as comprehensive documentation,
development and deployment practices; software licenses; career paths
for developers; identifiers for software through a
mechanism such as a Digital Object Identifier or publication of a ``software
paper''; and evidence of software contribution and impact, recorded in online
systems such as source code repositories like GitHub. For truly sustainable
software, there should be no endpoint, as the software products continue to be
used beyond a single institution, grant, and developer or
development team.

@@

Another problem Bourne experienced was staff retention from
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of software sustainability through public-private partnerships. He
noted that making a successful business from scientific software alone
is  rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding
requirement even with a vibrant community---goodwill only goes so
far if one is being paid to do something else.  He talked about grant
schemes of relevance in the U.S., particularly with regard to
technology transfer~\cite{sbir-web, fased-web}.

@@

The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon Smith of
GitHub. Smith started with an example from his past in data
reduction in Astronomy, where he needed to remove interfering effects from the
experimental apparatuses. He built a ``bad pixel mask,'' and realized that
while it was persistent, there was no way or practice of sharing these data
among scientists. Consequently many researchers repeated the same
calculations. Smith estimated that 13 person-years were wasted by this
repetition.

@@

C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor,
argued that with regard to reusable software, ``we should just start
doing it.''  Smith replied that documentation should be
``treated as a first class entity.''  He noted that the open source
community has excellent cultures of code reuse, for example,
RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
CPAN~\cite{cpan-web}, where there is effectively low-friction
collaboration through the use of repositories. This has not happened
in highly numerical, compiled language scientific software.  An
exception he cited as a good example of scientific projects using
GitHub is the EMCEE Markov Chain Monte Carlo project~\cite{emcee-web}
by Dan Foreman-Mackey and contributors.
